%% ID: three_particles
%% TITLE: Three Particles
%% TYPE: question
%% QUESTIONTYPE: numeric
%% CONCEPTS: momentum, energy, newtonii, momentumii, resolving_vectors, vectors, eq_of_motion_diff, relative_motion
%% VIDEOS: 
%% LEVEL: 5
%% TOPIC: mechanics/dynamics
%% ORDER: 10

\begin{problem}[A1988FMIVQ1a] % Requires vectors, v = dx/dt, s= integral(v)dx  relative velocities, a little parametric equations
{Three identical particles \vari{A}, \vari{B}, and \vari{C} are moving in a plane and, at time \vari{t}, their position vectors, \vari{\vtr{r_a}}, \vari{\vtr{r_b}}, \vari{\vtr{r_c}}, with respect to an origin \vari{O} are:
\begin{eqnarray*} 
\vtr{r_a} &=& (2t +1)\vtr{i} + (2t+ 3)\vtr{j}  \\ 
\vtr{r_b} &=& (10 - t)\vtr{i} + (12 - t)\vtr{j} \\ 
\vtr{r_c} &=& (t^{3} - 15t + 4)\vtr{i} + (-3t^{2} + 2t + 1)\vtr{j} 
\end{eqnarray*} 

\begin{enumerate}
	\item Find the magnitude of the velocity of particle \vari{C} relative to particle \vari{A} when \valuedef{t}{2}{}, and find the angle which this relative velocity makes with \vari{\vtr{i}} at this time.
	\item Verify that particles \vari{A} and \vari{B} are both moving along the straight line with equation \valuedef{y}{x + 2}{}, and that they collide when \valuedef{t}{3}{}.
	\item Given that the collision between particles \vari{A} and \vari{B} is elastic, find the velocities of \vari{A} and \vari{B} immediately after the collision.
	\item What are the position vectors of \vari{A} and \vari{B} a time \vari{\tau} after the collision?
\end{enumerate}
}
{\stress{Adapted with permission from UCLES, A Level Further Mathematics, Syllabus C, June~1988, Paper~4, Question~1.}}
{
\begin{enumerate}
	\item To find the relative velocity, we first need the velocities of \vari{A} and \vari{C}, which can be found by differentiating the position vectors (\vari{\vtr{s}_{A}} and \vari{\vtr{s}_{C}}) with respect to time:
\begin{eqnarray*} 
\vtr{v}_{A} &= \frac{\d \vtr{s}_{A}}{\d t} \\ 
&= \frac{\d}{\d t}\left[(2t + 1)\vtr{i} + (2t + 3)\vtr{j}\right] \\ 
&= 2\vtr{i} + 2\vtr{j} 
\end{eqnarray*}
where to differentiate the vector we have simply differentiated the two components separately. Repeating this for \vari{C}  gives:
\begin{eqnarray*} 
\vtr{v}_{C} & = \frac{\d \vtr{s}_{C}}{\d t} \\ 
&= (3t^{2} - 15)\vtr{i} + (6t + 2)\vtr{j} 
\end{eqnarray*}

The velocity of \vari{C} with respect to \vari{A} is simply \vari{\vtr{v}_{C} - \vtr{v}_{A}}:
\begin{eqnarray*} 
\vtr{v}_{\text{rel}} &= \vtr{v}_{C} - \vtr{v}_{A}\\ 
&= \left[(3t^{2} - 15)\vtr{i} + (-6t + 2)\vtr{j}\right] - \left[2\vtr{i} + 2\vtr{j}\right]\\ 
&= (3t^{2} - 17)\vtr{i} - (6t)\vtr{j} 
\end{eqnarray*}

At \valuedef{t}{2}{}, the relative velocity is \valuedef{\vtr{v}_{rel}(2)}{-5\vtr{i} - 12\vtr{j}}{}, which has a magnitude of \vari{v_{rel}} $=$ \valuedef{\sqrt{(-5)^{2} + (-12)^{2}}}{}{} $=$ \quantity{13}{m\,s\sup{-1}}. The angle the velocity makes to \vari{\vtr{i}} can be measured clockwise or anticlockwise, depending on convention. The vector makes an angle of \vari{\arctan\left(\frac{5}{12}\right)} to the negative \vari{x}-axis, which is $157.4^{\circ}$ clockwise or equivalently $202.6^{\circ}$ anticlockwise.
	\item The simplest way to show that \vari{A} and \vari{B} are travelling on the line \valuedef{y}{x + 2}{} is to show that the direction of motion is along the line \valuedef{y}{x}{} and that when \valuedef{x}{0}{} the \vari{y} coordinate is \vari{2}.

The direction of motion is the direction the velocity vector points in, provided there is no acceleration involved. We find \valuedef{vtr{v}_{B}}{-\vtr{i} - \vtr{j}}{} using the same process as before and, remembering that the \vari{\vtr{i}}-component is the \vari{x} and \vari{\vtr{j}}-component the \vari{y}, find also that the direction of motion is along \valuedef{y}{x}{} in each case. It is clear, however, from the fact that \vari{B} has negative components it is travelling along the line in the opposite direction.

To check the intercept, find the time that the \vari{x}-coordinate is zero for both particles; \valuedef{2t + 1}{0}{} and \valuedef{10 - t}{0}{} give \valuedef{t}{-\frac{1}{2}}{} for \vari{A} and \valuedef{t}{10}{} for \vari{B}. Substituting these times back in to the \vari{y}-coordinate for the particle concerned gives \valuedef{y}{2}{} at \valuedef{x}{0}{} for both: \valuedef{2\left(-\frac{1}{2}\right) + 3}{2}{} for \vari{A} and \valuedef{2 - (10)}{2}{} for \vari{B}. Hence both particles are moving along the line \valuedef{y}{x + 2}{}.

A more rigorous method would be to consider the position vectors as two parametric equations for \vari{x} and \vari{y}, then to find \vari{\frac{\d x}{\d t}} and \vari{\frac{\d y}{\d t}} then to use the Chain Rule to find
\begin{equation*} 
\frac{\d y}{\d x} = \frac{\left(\frac{\d y}{\d t}\right)}{\left( \frac{\d x}{\d t} \right)} = m
\end{equation*} 
for each particle. In both cases it comes out as \valuedef{m}{1}{}, and then we find the intercept, \vari{c}, as above to show the particles move along a line of the form \valuedef{y}{mx + c}{} where \valuedef{m}{1}{} and \valuedef{c}{2}{}.
	\item First we must find the time the collision occurs, and since we know they are moving on the same line in opposite directions; this must occur. Equating the \vari{x}-components of \vari{\vtr{v}_{A}} and \vari{\vtr{v}_{B}} and solving for \vari{t} gives the time at which the collision occurs:
\begin{eqnarray*} 
2t + 1 &= 10 -t \\ 
3t &= 9 \\ 
t &= 3 
\end{eqnarray*} Of course, in general we would need to check the \vari{y}-components were equal to show a collision occurs (the lines may cross, but the particles don't have to have the same \vari{y}-coordinate at the same time the \vari{x}-coordinates are the same); but here we know that it does.

Next we consider the conservation of momentum, which is a vector equation, and so we will leave it in vector form. Calling the components of the velocity of \vari{A} after the collision \vari{a} and \vari{\alpha}, and the components of \vari{B} after the collision \vari{b} and \vari{\beta}:
\begin{eqnarray*} 
(m)(2\vtr{i} + 2\vtr{j}) + (m)(-\vtr{i} - \vtr{j}) &= (m)(a\vtr{i} + \alpha\vtr{j}) + (m)(b\vtr{i} + \beta\vtr{j}) \\ 
\vtr{i} + \vtr{j} &= (a + b)\vtr{i} + (\alpha + \beta)\vtr{j}
\end{eqnarray*}
which produces two simultaneous equations, by considering the \vari{\vtr{i}} and \vari{\vtr{j}} components separately (they are linearly independent):
\begin{eqnarray} 
a + b &= 1 \label{CoM_vtr_i} \\ 
\alpha + \beta &= 1 \label{CoM_vtr_j} 
\end{eqnarray}

Then consider the conservation of energy, since the collision is elastic no energy is lost. Now we need the magnitude of the velocity squared: \vari{v^{2}} $=$ \valuedef{\vtr{v} \cdot \vtr{v}}{v_{x}^{2} + v_{y}^{2}}{}
\begin{eqnarray} 
\frac{1}{2}mv_{A}^{2} + \frac{1}{2}mv_{B}^{2} &= \frac{1}{2}mv_{A_{f}} + \frac{1}{2}mv_{B_{f}} \notag \\ 
\left(2^{2} + 2^{2}\right) + \left((-1)^{2} + (-1)^{2}\right) &= \left(a^{2} + \alpha^{2}\right) + \left(b^{2} + \beta^{2}\right) \notag \\ 
10 &= a^{2} + \alpha^{2} + b^{2} + \beta^{2} \label{CoE_vtr} 
\end{eqnarray}
where the subscript \vari{f} denotes after the collision.

This gives us three simultaneous equations, \eqref{CoM_vtr_i}, \eqref{CoM_vtr_j} and \eqref{CoE_vtr}, for four unknowns; we need another equation in order to solve them. We know this is a 1D collision; the particles move along the line \valuedef{y}{x + 2}{}, and so we know the ratio of the \vari{x} and \vari{y} components of the velocities must be the gradient, equal to \vari{1}: 
\begin{eqnarray} 
\frac{a}{\alpha} &= 1 \label{1D_vtr_x} \\ 
\frac{b}{\beta} &= 1 \label{1D_vtr_y} 
\end{eqnarray}
It now looks like we have 5 equations for 4 unknowns. If they are 5 independent equations, then the solution is overdetermined and we cannot solve for it; fortunately here they are not. Equating the left hand sides of Equations \eqref{CoM_vtr_i} and \eqref{CoM_vtr_j}, then using Equation \eqref{1D_vtr_x} to remove \vari{a} and \vari{b} and finally dividing by \vari{\beta} gives Equation \eqref{1D_vtr_y}. So Equation \eqref{1D_vtr_y} can be derived from the others; it is not independent and we can ignore it when solving the simultaneous equations (or in fact use it and neglect any of the others, except \eqref{CoE_vtr} which must be used).

Solving  Equations \eqref{CoM_vtr_i}, \eqref{CoM_vtr_j}, \eqref{CoE_vtr} and \eqref{1D_vtr_x} for the 4 unknowns produces two solutions: the initial conditions, \vari{a} $=$ \valuedef{\alpha}{2}{} and \vari{b} $=$ \valuedef{\beta}{-1}{} which correspond to \vari{\vtr{v}_{A}} and \vari{\vtr{v}_{B}} but also the final velocities we are looking for. Writing them in vector form:
\begin{eqnarray*} 
\vtr{v}_{A_{f}} &= - \vtr{i}  - \vtr{j} \\ 
\vtr{v}_{B_{f}} &= 2\vtr{i} + 2\vtr{j} 
\end{eqnarray*}
which is simply a swapping over of velocities, as expected in an elastic collision with identical particles.
	\item The final part of the question is simply integrating up the two velocities, but with respect to \vari{\tau} instead of \vari{t} because \valuedef{t}{3}{} when the collision occurs and we want the variable equal to zero at that time. That trick aside, the question is a straightforward indefinite integration:
\begin{eqnarray*}
 \vtr{s}_{A_{f}} &= \int \vtr{v}_{A_{f}} \d \tau \\ 
 &= \int (-1)\vtr{i} + (-1)\vtr{j} \d \tau \\
  &= \left( \int [-1] \d \tau \right)\vtr{i} + \left( \int [-1] \d \tau \right) \\ 
  &= (-\tau + c_{1})\vtr{i} + (-\tau + c_{2})\vtr{j} 
  \end{eqnarray*} 
 where \vari{c_{1}} and \vari{c_{2}} are arbitrary constants of integration, and the two parts of the vector are simply integrated separately. They can be found by equating the position vector \vari{\vtr{s}_{A_{f}}} at \valuedef{\tau}{0}{} with the position of the collision, \vari{\vtr{s}_{A}} at \valuedef{t}{3}{} which is \vari{7\vtr{i} + 9\vtr{j}}. This gives \valuedef{c_{1}}{7}{} and \valuedef{c_{2}}{9}{}.

Repeating this for particle \vari{B}, we find the position vectors at a time \vari{\tau} after the collision:
\begin{eqnarray*}
 \vtr{s}_{A_{f}} &= (7 - \tau)\vtr{i} + (9 - \tau)\vtr{j} \\ 
 \vtr{s}_{B_{f}} &= (2\tau + 7)\vtr{i} + (2\tau + 9)\vtr{j} 
 \end{eqnarray*}
\end{enumerate}
}
\end{problem}